Follow-up Samples

### 2017-02-27 submission/manuscript_outline_20161024.tex

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
Specifically, we wanted to identify components within the community that
were different before and after removal of said lesion.

@@

\textbf{Conclusions:} Our data supports the hypothesis that there are
changes in the bacterial microbiome following colorectal cancer lesion
removal. Individuals with carcinoma have more drastic differences to the
overall community then those with adenoma. Changes to commensal bacteria
were some of the most important variables for model classification,
suggesting that these bacteria may be central to initial polyp formation
and transition to carcinoma.

@@

Over the last few years studies have shown how the bacterial microbiome
{[}3{]} or specific members within it {[}4{]} could be directly involved
with the pathogenesis of CRC. There has also been research into how the
bacterial microbiome could be altered directly on tumor tissue itself
{[}5{]}. These studies have helped to provide a tantalizing link between
the bacterial microbiome and CRC. Building off of these findings, there
has been promising work on the bacterial microbiome and its ability to
complement existing screening methods such as Fecal Immunoglobulin Test
(FIT) or act alone as a screening tool {[}6,7{]}. Although these studies
suggest that the bacterial microbiome might change after treatment there
remains limited information on the bacterial microbiome response to
lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as the relative abundance of specific Operational Taxonomic
Units (OTUs). We utilized Random Forest to build two models. The first
model was built to classify lesion and normal (non-lesion) samples while
the second was built to classify initial and follow up samples.
Subsequent observations on how these models, as well as specific OTUs
within them, performed pre and post lesion removal helped inform us as
to whether initial and follow up samples were changing and whether it
was towards a more normal bacterial microbiome or not. We also
investigated the two models for similar important OTUs to identify the
crucial OTUs for not only classifying lesion or normal samples but also
initial and follow up.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs, and investigating the
relative abundance of specific previously associated CRC bacteria
(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}) based on adenoma and carcinoma. From here the lesion model
was then tested for accuracy in prediction and whether it reduced the
positive probability of lesion after surgery. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model and whether it could reduce the positive probability of an
initial sample in the follow up samples. Finally, a list of common OTUs
were found for the two different models used.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after lesion (i.e.~adenoma and carcinoma) removal.
Specifically, we wanted to identify components within the community that
were different before and after removal of said lesion.

@@

\textbf{Conclusions:} Our data supports the hypothesis that there are
changes in the bacterial microbiome following colorectal lesion removal.
Individuals with carcinoma have more drastic differences to the overall
community then those with adenoma. Changes to commensal bacteria were
some of the most important variables for model classification,
suggesting that these bacteria may be central to initial polyp formation
and transition to carcinoma.

@@

Over the last few years studies have shown how the bacterial microbiome
{[}3{]} or specific members within it {[}4{]} could be directly involved
with the pathogenesis of CRC. There has also been research into how the
bacterial microbiome could be altered directly on tumor tissue itself
{[}5{]}. These studies have helped to provide a tantalizing link between
the bacterial microbiome and CRC. Building off of these findings, there
has been promising work on the bacterial microbiome and its ability to
complement existing screening methods such as Fecal Immunoglobulin Test
(FIT) or act alone as a screening tool {[}6,7{]}. Although these studies
suggest that the bacterial microbiome might change after treatment there
remains limited information on the bacterial microbiome response to
lesion (adenoma and carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion. We
analyzed changes in alpha and beta diversity as well as the relative
abundance of specific Operational Taxonomic Units (OTUs). We utilized
Random Forest to build two models. The first model was built to classify
lesion versus non-lesion (normal) samples while the second was built to
classify initial versus follow up samples. Subsequent observations on
how these models, as well as specific OTUs within them, performed pre
and post lesion removal helped inform us as to whether initial and
follow up samples were changing and whether it was towards a more normal
bacterial microbiome or not. We also investigated the two models for
similar important OTUs to identify the crucial OTUs for not only
classifying lesion or normal samples but also initial versus follow up.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after removal {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after lesion removal {[}Figure
2{]}.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs, and investigating the
relative abundance of specific previously associated CRC bacteria
(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}) based on adenoma or carcinoma. From here the lesion model was
then tested for accuracy in prediction and whether it reduced the
positive probability of lesion after surgery. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model and whether it could reduce the positive probability of an
initial sample in the follow up samples. Finally, a list of common OTUs
were found for the two different models used.

### 2017-02-27 submission/supplemental_outline_20161024.Rmd

loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))

@@

st2 <- select(common_vars_summary, otu) %>% 
  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
  mutate(Pvalue = common_vars_summary$Pvalue) %>% 
  mutate(BH_corr = common_vars_summary$BH_corr)

colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")

>>>>

loadLibs(c("tidyr", "dplyr", "scales", "knitr", "rmarkdown"))

@@

lesion_top10_varsMDA <- read.csv('../results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
IF_top10_varsMDA <- read.csv('../results/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)

# Create selection vector
common_vars <- common_vars_summary$otu

# Create MDA vectors
lesion_MDA <- filter(lesion_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)

initial_MDA <- filter(IF_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)

st2 <- select(common_vars_summary, otu) %>% 
  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
  mutate(lesion_MDA = lesion_MDA$mean_MDA) %>% 
  mutate(initial_MDA = initial_MDA$mean_MDA) %>% arrange(desc(lesion_MDA))

colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Lesion Model MDA", "Initial Model MDA")
